---
output:
  pdf_document:
    fig_caption: true
    number_sections: false
    latex_engine: pdflatex
    fig_crop: false
  word_document:
    fig_caption: true
    number_sections: false
fontsize: 12pt
geometry: margin=1in
header-includes:
  - \usepackage{setspace}
  - \usepackage{newtxtext,newtxmath}
  - \singlespacing
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{float}
  - \usepackage{pdflscape}
  - \usepackage{fvextra}
  - \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,breakanywhere=true,fontsize=\scriptsize,commandchars=\\\{\}}
  - \setlength{\parindent}{0pt}
  - \setlength{\parskip}{0.55em}
  - \raggedbottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "90%"
)

pkg = c("knitr", "dplyr", "tidyr", "tibble", "ggplot2", "jpeg", "png", "flextable", "kableExtra")
need = pkg[!vapply(pkg, requireNamespace, logical(1), quietly = TRUE)]
if(length(need) > 0) install.packages(need, repos = "https://cloud.r-project.org")
invisible(lapply(pkg, library, character.only = TRUE))
library(grid)

pagebreak = function(){
  cat("\n\\newpage\n")
}

# Edit this path once on your own computer if needed.
dataPath = "D:/machine learning/final project/rscbjbr9sj-2/ChestXRay2017/chest_xray"

findDataDir = function(path = dataPath){
  inputFile = tryCatch(knitr::current_input(), error = function(e) "")
  inputDir = if(is.null(inputFile) || identical(inputFile, "")) getwd() else dirname(normalizePath(inputFile, winslash = "/", mustWork = FALSE))
  candidates = unique(c(
    path,
    file.path(getwd(), "chest_xray"),
    file.path(inputDir, "chest_xray")
  ))
  hit = candidates[dir.exists(candidates)]
  if(length(hit) == 0) return(NA_character_)
  normalizePath(hit[1], winslash = "/", mustWork = FALSE)
}

cxrDir = findDataDir()

splitTab = tibble(
  Dataset = c("test", "train"),
  Count = c(624, 5232),
  Percentage = c("10.66%", "89.34%")
)

threeClassTab = tibble(
  Dataset = c("test", "test", "test", "train", "train", "train"),
  Sublabel = c("BACTERIA", "NORMAL", "VIRUS", "BACTERIA", "NORMAL", "VIRUS"),
  Count = c(242, 234, 148, 2538, 1349, 1345),
  Percentage = c("38.78%", "37.50%", "23.72%", "48.51%", "25.78%", "25.71%")
)

geomTab = tibble(
  Dataset = c("test", "test", "test", "train", "train", "train"),
  Sublabel = c("BACTERIA", "NORMAL", "VIRUS", "BACTERIA", "NORMAL", "VIRUS"),
  Count = c(5, 3, 5, 42, 24, 21),
  WidthMean = c(1211, 1652, 1489, 1125, 1666, 1308),
  HeightMean = c(812, 1361, 1180, 777, 1378, 887),
  PixelsMean = c(1024618, 2284435, 1781595, 940366, 2360966, 1215628),
  AspectMean = c(1.49, 1.22, 1.27, 1.53, 1.22, 1.54)
)

issueTab = tibble(
  Finding = c(
    "Class imbalance",
    "Patient leakage",
    "Image geometry differences",
    "Inconsistent image format",
    "Train-test intensity differences",
    "Very small validation set",
    "Class signal in basic image features",
    "Limited benefit from class weighting"
  ),
  Evidence = c(
    "Pneumonia images account for 73.0% of the dataset.",
    "172 of 204 reconstructed test patients also appear in training.",
    "NORMAL images tend to be wider and closer to square in the sampled audit.",
    "Most images are grayscale, but 283 files are RGB.",
    "Pixel-intensity summaries differ between the training and test sets.",
    "The exploratory validation split contains only 16 images.",
    "Basic image features from the original files separate classes better than expected by chance.",
    "Balanced class weights changed macro-F1 only slightly."
  ),
  Action = c(
    "Report macro-F1, balanced accuracy, and class-specific results instead of overall accuracy alone.",
    "Rebuild the data split at the patient level before interpreting model performance.",
    "Preserve aspect ratio during preprocessing and test whether geometry still predicts class.",
    "Convert all images to a single grayscale format before modeling.",
    "Normalize after splitting and re-check intensity distributions.",
    "Use a larger validation set or grouped cross-validation for model selection.",
    "Compare image models with simple baselines built on basic file-level features.",
    "Combine class weighting with label review and error analysis."
  )
)

makeTable = function(df, caption, align = NULL, font_size = 10){
  if(knitr::pandoc_to() == "docx" && requireNamespace("flextable", quietly = TRUE)){
    ft = flextable::flextable(df)
    ft = flextable::set_caption(ft, caption = caption)
    ft = flextable::font(ft, fontname = "Times New Roman", part = "all")
    ft = flextable::fontsize(ft, size = font_size, part = "all")
    ft = flextable::align(ft, align = "center", part = "header")
    ft = flextable::align(ft, align = "left", part = "body")
    ft = flextable::valign(ft, valign = "center", part = "all")
    ft = flextable::padding(ft, padding.top = 5, padding.bottom = 5, padding.left = 6, padding.right = 6, part = "all")
    ft = flextable::line_spacing(ft, space = 1.10, part = "all")
    ft = flextable::set_table_properties(ft, layout = "autofit", width = 0.82)
    return(ft)
  }

  out = knitr::kable(
    df,
    caption = caption,
    align = align,
    booktabs = knitr::is_latex_output(),
    linesep = "",
    longtable = FALSE
  )
  if(knitr::is_latex_output() && requireNamespace("kableExtra", quietly = TRUE)){
    out = kableExtra::kable_styling(
      out,
      latex_options = c("hold_position"),
      position = "center",
      font_size = font_size
    ) |>
      kableExtra::row_spec(0, extra_latex_after = "\\addlinespace[0.10cm]") |>
      kableExtra::row_spec(1:nrow(df), extra_latex_after = "\\addlinespace[0.06cm]")
  }
  out
}

renderTable4 = function(df, caption){
  if(knitr::is_latex_output() && requireNamespace("kableExtra", quietly = TRUE)){
    out = knitr::kable(
      df,
      caption = caption,
      align = c("l", "l", "l"),
      booktabs = TRUE,
      linesep = "",
      longtable = FALSE
    ) |>
      kableExtra::kable_styling(
        latex_options = c("hold_position"),
        position = "center",
        font_size = 9
      ) |>
      kableExtra::column_spec(1, width = "1.10in") |>
      kableExtra::column_spec(2, width = "3.65in") |>
      kableExtra::column_spec(3, width = "3.65in") |>
      kableExtra::row_spec(0, extra_latex_after = "\\addlinespace[0.16cm]") |>
      kableExtra::row_spec(1:nrow(df), extra_latex_after = "\\addlinespace[0.14cm]") |>
      kableExtra::landscape()
    return(knitr::asis_output(as.character(out)))
  }

  if(knitr::pandoc_to() == "docx" && requireNamespace("flextable", quietly = TRUE)){
    ft = flextable::flextable(df)
    ft = flextable::set_caption(ft, caption = caption)
    ft = flextable::font(ft, fontname = "Times New Roman", part = "all")
    ft = flextable::fontsize(ft, size = 9.5, part = "all")
    ft = flextable::width(ft, j = 1, width = 1.15)
    ft = flextable::width(ft, j = 2, width = 3.45)
    ft = flextable::width(ft, j = 3, width = 3.45)
    ft = flextable::align(ft, align = "left", part = "all")
    ft = flextable::valign(ft, valign = "top", part = "all")
    ft = flextable::padding(ft, padding.top = 6, padding.bottom = 6, padding.left = 7, padding.right = 7, part = "all")
    ft = flextable::line_spacing(ft, space = 1.18, part = "all")
    ft = flextable::set_table_properties(ft, layout = "fixed", width = 0.98)
    return(ft)
  }

  makeTable(df, caption, align = c("l", "l", "l"), font_size = 9)
}

readImageAny = function(path){
  ext = tolower(tools::file_ext(path))
  if(ext %in% c("jpg", "jpeg")) jpeg::readJPEG(path) else png::readPNG(path)
}

pickExample = function(split, label){
  if(is.na(cxrDir)) return(NA_character_)
  if(label == "NORMAL"){
    folder = file.path(cxrDir, split, "NORMAL")
    files = list.files(folder, pattern = "\\.(jpeg|jpg|png)$", full.names = TRUE, ignore.case = TRUE)
  } else {
    folder = file.path(cxrDir, split, "PNEUMONIA")
    files = list.files(folder, pattern = paste0(label, ".*\\.(jpeg|jpg|png)$"), full.names = TRUE, ignore.case = TRUE)
  }
  if(length(files) == 0) return(NA_character_)
  files[1]
}
```

```{r titlepage, results='asis'}
if(knitr::is_latex_output()){
  cat("\\begin{titlepage}\n")
  cat("\\thispagestyle{plain}\n")
  cat("\\centering\n")
  cat("\\vspace*{1.55in}\n")
  cat("{\\LARGE Classification of Chest X-Ray Using Convolutional\\par}\n")
  cat("{\\LARGE Networking\\par}\n")
  cat("\\vspace{0.55in}\n")
  cat("{\\large Hanjun Ni(whs9hm), Yifeng Song(ys8mz), Grace Zhou(jgg8xd)\\par}\n")
  cat("\\vspace{0.30in}\n")
  cat("{\\large 2026-03-08\\par}\n")
  cat("\\vfill\n")
  cat("\\end{titlepage}\n")
} else {
  cat("<div style='text-align:center; margin-top:180px;'>\n")
  cat("<p style='font-size:22pt; margin-bottom:28px;'>Classification of Chest X-Ray Using Convolutional<br>Neural Networking</p>\n")
  cat("<p style='font-size:14pt; margin-bottom:18px;'>Hanjun Ni(whs9hm), Yifeng Song(ys8mz), Grace Zhou(jgg8xd)</p>\n")
  cat("<p style='font-size:14pt;'>2026-03-08</p>\n")
  cat("</div>\n\n")
  pagebreak()
}
```
## Hanjun Ni(whs9hm), Yifeng Song(ys8mz), Grace Zhou(jgg8xd)
## Background
The data set that was utilized for conducting this research was retrieved from Mendeley Data and corresponds to the chest X-ray collection put together by Kermany, Zhang, and Goldbaum. The public files are large enough and are set up in both a training folder and a testing folder, making it a suitable choice for a class project in the field of image classification. This collection was established as a supplement to the research article “Identifying Medical Diagnoses and Treatable Diseases by Image-Based Deep Learning,” published in the journal Cell. 

The primary purpose for collecting and validating these thousands of images was to develop and test deep learning frameworks capable of identifying medical diagnoses through automated image analysis. Scientifically, this dataset provides a large-scale, validated repository for advancing research in medicine and ophthalmology. Socially, the data supports the development of diagnostic tools that could lead to faster identification of treatable diseases, potentially improving clinical efficiency and patient outcomes. It is essential that the initial split of the data set is evaluated with due diligence. An accurate chest X-ray classifier should identify patterns in the X-ray image that represent a disease; however, it should not be influenced by artifacts caused by the way the folders and filenames were created, the way the images were captured, or by the manner of how the images are acquired.

Thus, prior to modelling using a Convolution Neural Network (CNN), this project will perform a statistical review of the dataset. The intent is to assess if the apparent predictions are driven exclusively by pathology or if they are also influenced by shortcut learning found in the data set. This distinction is relevant because a model demonstrating excellent predictive accuracy does not necessarily demonstrate clinically meaningful learning. This is especially true where it is possible that the patients from the data set may exhibit overlapping patients amongst the train and test splits; there may be mixed image modalities present within the data set; there may be an imbalance in the geometry of the images; and there may be inherent differences in pixel intensities between the training and testing images.



## Summary of Data

There are 5,856 chest X-ray images in the dataset. There are 5,232 images for training and 624 images for testing, as indicated in (Table 1). The images in the dataset can be divided into three categories according to their folder structure and the naming convention of the files. These categories are NORMAL, BACTERIA, and VIRUS. The frequency of each category is presented in both (Table 2) and (Figure 2). In both sets of data, the majority of cases of pneumonia fall into the pneumonia image group. As a result, when evaluating the model later in the report, it is important to consider metrics beyond just raw accuracy.
(Figure 1) contains a representative example of images contained in the dataset.

The results of the examination of the files also reveal evidence of geometric neutrality among the files. The summary indicated that the images which fall into the NORMAL category tend to be of larger size and more square than those images which fall into either the bacterial pneumonia or the viral pneumonia categories. In contrast, the images which fall into either the bacterial pneumonia or the viral pneumonia category tend to be of smaller size and exhibit greater variation in aspect ratio than the images which fall into the NORMAL category (as illustrated in Table 3 & Figure 3). Other concerns were discovered through more in-depth research. These concerns included: differing image colour modes; train and test partitioning with notable brightness difference; extremely small partitioning of images for validation; the presence of extreme collinearity among width, height, and pixel count; and significant overlap between patients in the training and test portions of the datasets. The major results of the preliminary research were summarized in (Table 4). The results of the patient overlap concern are shown in (Figure 4).

## Potential Problem and Future Plan

The primary issue being addressed in this paragraph is the determination of whether a classifier trained using the original split of data will learn true disease-related patterns or will instead take advantage of features present in the dataset. So far there have been a number of known risks to consider, including: overlap between patients; class imbalances; geometric and intensity-based characteristics; differences in color modes; possible sub-typing label ‘noise’; and minimal stability of sensitivity analysis due to an inadequate size of the validation sample.

For these reasons, future phases of this effort need to focus on three areas of priority: create convincing splits based on patient level; ensure consistent pre-processing methods across all cases; and use similar methods for grouped evaluations. As such, CNN results cannot be construed until after addressing each of these aforementioned issues.

## Reference

Dablain, D., Jacobson, K. N., Bellinger, C. et al. *Understanding CNN Fragility When Learning with Imbalanced Data*. Machine Learning, 113, 4785-4810 (2024).

Kermany, D., Zhang, K., and Goldbaum, M. *Large Dataset of Labeled Optical Coherence Tomography (OCT) and Chest X-Ray Images*. Mendeley Data, V3, doi:10.17632/rscbjbr9sj.3.

\newpage

# Tables and Figures

```{r table1, results='asis'}
makeTable(splitTab, "Summary of split data.", align = c("l", "r", "r"), font_size = 11)
```

\vspace{0.65em}

\newpage

```{r table2, results='asis'}
makeTable(threeClassTab, "Class frequency by split and subtype.", align = c("l", "l", "r", "r"), font_size = 11)
```

\vspace{0.65em}

\newpage

```{r table3, results='asis'}
makeTable(geomTab, "Image dimensions summary by category.", align = c("l", "l", "r", "r", "r", "r", "r", "r"), font_size = 9)
```

```{r table4, results='asis'}
renderTable4(issueTab, "Key findings and modeling actions.")
```

\newpage

```{r fig1, fig.cap="Representative chest X-ray images by class and split.", fig.width=8.5, fig.height=5.5, out.width='95%'}
layout = expand.grid(split = c("train", "test"), label = c("NORMAL", "BACTERIA", "VIRUS"), stringsAsFactors = FALSE)
layout = layout[c(1, 3, 5, 2, 4, 6), ]
paths = mapply(pickExample, layout$split, layout$label, USE.NAMES = FALSE)

grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 3)))
for(i in seq_len(nrow(layout))){
  r = ifelse(i <= 3, 1, 2)
  c = ifelse(i <= 3, i, i - 3)
  vp = viewport(layout.pos.row = r, layout.pos.col = c)
  pushViewport(vp)
  grid.rect(gp = gpar(col = "grey70", fill = "white"))
  if(!is.na(paths[i]) && file.exists(paths[i])){
    img = readImageAny(paths[i])
    grid.raster(img, width = unit(0.95, "npc"), height = unit(0.82, "npc"), interpolate = FALSE)
  } else {
    grid.text("Image not found", y = 0.55, gp = gpar(cex = 0.95, fontfamily = "serif"))
    grid.text("Update dataPath in setup", y = 0.40, gp = gpar(cex = 0.80, col = "grey35", fontfamily = "serif"))
  }
  grid.text(paste(toupper(layout$split[i]), layout$label[i], sep = " - "),
            y = unit(0.96, "npc"), gp = gpar(cex = 0.78, fontface = "plain", fontfamily = "serif"))
  popViewport()
}
```

\newpage

```{r fig2, fig.cap="Three-class distribution in the train and test splits.", fig.width=7.4, fig.height=4.6, out.width='88%'}
fig2dat = threeClassTab |>
  mutate(Sublabel = factor(Sublabel, levels = c("NORMAL", "BACTERIA", "VIRUS")),
         Dataset = factor(Dataset, levels = c("train", "test")))

ggplot(fig2dat, aes(x = Dataset, y = Count, fill = Sublabel)) +
  geom_col(position = "dodge", width = 0.65, colour = "black", linewidth = 0.2) +
  labs(x = NULL, y = "Count") +
  theme_minimal(base_size = 11, base_family = "serif") +
  theme(
    legend.title = element_blank(),
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.title.y = element_text(size = 10),
    plot.margin = margin(5.5, 12, 5.5, 5.5)
  )
```

\newpage

```{r fig3, fig.cap="Mean width and mean aspect ratio in the sampled geometry audit.", fig.width=7.6, fig.height=5.2, out.width='88%'}
par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.2, 1.2), oma = c(0, 0, 0, 0), family = "serif")
cols = c("grey80", "grey55", "grey35")
barplot(height = geomTab$WidthMean,
        names.arg = paste(substr(geomTab$Dataset, 1, 1), substr(geomTab$Sublabel, 1, 3), sep = "-"),
        col = cols[match(geomTab$Sublabel, c("NORMAL", "BACTERIA", "VIRUS"))],
        border = "grey20",
        las = 2,
        ylab = "Mean width")
barplot(height = geomTab$AspectMean,
        names.arg = paste(substr(geomTab$Dataset, 1, 1), substr(geomTab$Sublabel, 1, 3), sep = "-"),
        col = cols[match(geomTab$Sublabel, c("NORMAL", "BACTERIA", "VIRUS"))],
        border = "grey20",
        las = 2,
        ylab = "Mean aspect ratio")
par(mfrow = c(1, 1))
```

\newpage

```{r fig4, fig.cap="Reconstructed patient overlap among the test cases.", fig.width=6.0, fig.height=4.4, out.width='72%'}
leakDat = tibble(
  Group = factor(c("Overlap with train", "Unique to test"), levels = c("Overlap with train", "Unique to test")),
  Count = c(172, 32)
)

ggplot(leakDat, aes(x = Group, y = Count)) +
  geom_col(width = 0.58, fill = "grey60", colour = "black", linewidth = 0.25) +
  geom_text(aes(label = Count), vjust = -0.5, size = 3.8, family = "serif") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = NULL, y = "Number of test patients") +
  theme_minimal(base_size = 11, base_family = "serif") +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(size = 10),
    plot.margin = margin(5.5, 16, 5.5, 16)
  )
```

\newpage

# Appendix

The appendix gives a compact record of how the proposal tables and figures were produced. The code below keeps only the core steps needed to reconstruct the audit from the original notebooks.

## A1. Build image-level metadata

```{r appendix1, echo=TRUE, eval=FALSE}
library(dplyr)
library(purrr)
library(jpeg)
library(png)

baseDir = "D:/machine learning/final project/rscbjbr9sj-2/ChestXRay2017/chest_xray"

extractPatient = function(path, label){
  name = basename(path)
  if(label == "NORMAL") return(sub("-[0-9]{4}.*$", "", name))
  hit = regmatches(name, regexpr("person[0-9]+", name))
  if(length(hit) == 0 || hit == "") return(name)
  hit
}

readGrayInfo = function(path){
  ext = tolower(tools::file_ext(path))
  img = if(ext %in% c("jpg", "jpeg")) jpeg::readJPEG(path) else png::readPNG(path)
  d = dim(img)
  if(length(d) == 2){
    gray = img
    mode = "L"
    h = d[1]
    w = d[2]
  } else {
    gray = 0.2989 * img[, , 1] + 0.5870 * img[, , 2] + 0.1140 * img[, , 3]
    mode = ifelse(d[3] == 3, "RGB", paste0("C", d[3]))
    h = d[1]
    w = d[2]
  }
  tibble(width = w, height = h, pixels = w * h, ratio = w / h,
         mean = mean(gray) * 255, sd = sd(as.numeric(gray)) * 255, mode = mode)
}
```

## A2. Produce the descriptive summary tables

```{r appendix2, echo=TRUE, eval=FALSE}
splitTab = meta |>
  count(dataset) |>
  mutate(Percentage = paste0(round(100 * n / sum(n), 2), "%"))

threeClassTab = meta |>
  count(dataset, sublabel) |>
  group_by(dataset) |>
  mutate(Percentage = paste0(round(100 * n / sum(n), 2), "%")) |>
  ungroup()
```

## A3. Reconstruct overlap and audit shortcut risk

```{r appendix3, echo=TRUE, eval=FALSE}
trainPatients = unique(meta$patientid[meta$dataset == "train"])
testPatients  = unique(meta$patientid[meta$dataset == "test"])
overlapPatients = intersect(trainPatients, testPatients)
leakageRate = round(100 * length(overlapPatients) / length(testPatients), 1)

modeTab = table(meta$mode)
cor(meta[, c("width", "height", "pixels", "ratio", "mean", "sd")], use = "pairwise.complete.obs")
```

## A4. Generate the figures used in the proposal

```{r appendix4, echo=TRUE, eval=FALSE}
# Figure 1: sample image panel
# Figure 2: class distribution by split
# Figure 3: width and aspect-ratio audit
# Figure 4: patient overlap summary
```
